---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer: It gives a blank plot, the data is loaded but we have not asked it to do anything with the data.

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:It adds an axis, has the data, but it does not have any geom points.

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point()
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point(aes(colour=cut))
  
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(aes(colour=cut)) +
  geom_smooth()

```

- Assign that last plot to an object called `obds_diamonds`.

```{r}

obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(aes(colour=cut)) +
  geom_smooth()


```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer: aesthetic changes at the start are transferred later - so 1 is divided by colour from the start, whereas 2 is not.

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
data(ChickWeight)
str(ChickWeight)
```

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight)) +
  geom_point()
```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point()

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point() +
  geom_smooth(method = lm)

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:diet 3

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~Diet)


```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~Diet)



```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
#this is a special way of loading the data which is packaged in ggplot2.
ggplot2::msleep
str(msleep)
```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x=order)) +
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = 4, hjust = 1, angle = 90))

#hjust is how you align text: 1 is right, 0 is left, 0.5 is middle.

```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x=order)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = 6, hjust = 1, angle = 90),
        axis.text.y = element_text(size = 15))

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}
str(msleep)
ggplot(msleep, aes(x=order, fill=genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = 6, hjust = 1, angle = 90),
        axis.text.y = element_text(size = 15))

```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}

str(msleep)
ggplot(msleep, aes(x=order, fill=genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = 2, hjust = 1, angle = 90),
        axis.text.y = element_text(size = 2),
        legend.key.size = unit(1, 'mm')
        )

```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}

str(msleep)
ggplot(msleep, aes(x=order, fill=genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = 6, hjust = 1, angle = 90),
        axis.text.y = element_text(size = 6),
        legend.key.size = unit(1, 'mm'),
        ) +
  guides(fill=guide_legend(ncol=3))

```

- Add a contour of thin black lines to the bars.

```{r}
ggplot(msleep, aes(x=order, fill=genus)) +
  geom_bar(colour="black", linewidth=0.1) +
  theme(axis.text.x = element_text(size = 6, hjust = 1, angle = 90),
        axis.text.y = element_text(size = 6),
        legend.key.size = unit(1, 'mm'),
        ) +
  guides(fill=guide_legend(ncol=3))

```

- Assign that last plot to an object called `obds_msleep`.

```{r}

obds_msleep <-ggplot(msleep, aes(x=order, fill=genus)) +
  geom_bar(colour="black", linewidth=0.1) +
  theme(axis.text.x = element_text(size = 6, hjust = 1, angle = 90),
        axis.text.y = element_text(size = 6),
        legend.key.size = unit(1, 'mm'),
        ) +
  guides(fill=guide_legend(ncol=3))
# as you have filled by genus, you select fill=guide
# so if you have done colour in aes at the very top, you would do colour=guide_legend... etc. 

obds_msleep

```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=12, fig.width=12}
library(cowplot)
first_row <- cowplot::plot_grid(obds_diamonds, labels = c("A")
  
  
)
second_row <- cowplot::plot_grid(obds_chickweight, obds_msleep, labels = c("B", "C")
  
)
super_plot <- cowplot::plot_grid(first_row, second_row, ncol =1, nrow=2
  
  
)
super_plot
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)
(obds_diamonds) / (obds_chickweight + obds_msleep)

```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)

gridExtra::grid.arrange(obds_diamonds, obds_msleep, ncol = 2, nrow = 1)

#nb can't save easily

```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave(filename = "super_plot.png", plot = super_plot, width=20, height=20, units = "cm", dpi = 70)
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}




```
